Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-04 paper/Intuition.tex
Intuition cleanup

In this section, we present the proof structure for the soundness property of our
MSI-based cache coherence protocol of Figure \ref{trans}. Specifically, we
formally reduce the proof of the store atomicity theorem for the whole system
to the proof of a set of invariants. In order to prove these invariants we
introduce the main ``intuition'' behind the correctness of the protocol. This
intuition is not restricted to the MSI-based protocol; Section
\ref{sec:Refinements} shows that it applies to other protocols like MOSI.

@@

We've defined store atomicity to be a whole system property. The first step in
our framework is to reduce the store atomicity property down to the following 4
global invariants about nodes:

\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ iff
the state of the cache for that address is at least \Sh{} and the
state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request has been
  processed at any time $t_i$ such that $t_i < t$, or
\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

We use the definition of compatibility between two caches as follows:
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible iff
either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
\label{compatible}
\end{defn}

@@

We can see that the correctness of our system (Theorem~\textit{storeAtomicity})
follows naturally from Invariants \ref{latestValue} to \ref{noReqAgain}.  A
store for an address may happen only when leaf cache is in \Mo{} state for that
address (Invariant~\textit{processReq}). Any two distinct leaf caches are not
ancestors of each other, so by Invariant \textit{nonAncestorCompatible}, if a
leaf cache is in \Mo{} state, all other leaf caches will be in \In{} state
which prevents processing of both load and store requests for the same address
in other leaf caches (again by Invariant \textit{processReq}). Thus, Definition
\textit{uniqStTimes} holds.  Given Invariant~\textit{noReqAgain}, once a
response has been produced for a request $i$, the cache may subsequently only
process $j^{th}$ request such that $j > i$. Thus there is at most one response
to each request, satisfying Definition~\textit{uniqRespLabels}. A load request
can be processed only if a cache is at least in \Sh{} state (again by Invariant
\textit{processReq}). When a leaf cache is at least in \Sh{} state, since it
has no children, it is \clean. A load response supplies the data for that
address present in the cache, which, by Invariant~\textit{lastestValue},
contains the ``latest value''. Thus, the final condition for
Theorem~\textit{storeAtomicity} is also satisfied.

@@

Invariants~\textit{processReq} and \textit{noReqAgain} can be proved directly
from the atomic transitions describing the system (Figure~\ref{trans} for MSI;
other protocols must also have similar transitions). Proving
invariants~\textit{uniqM} and \textit{latestValue} needs further work. We can
prove both using the following invariants on parent-child relationships.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in the
cache tree hierarchy, from Invariants \textit{conservative} and the first part
of \textit{localCompatibility}, we can show that for any address address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor relation is the
reflective-transitive closure of the parent relation, and Invariants
\textit{conservative} and the first part \textit{localCompatibility} of defines
a $\ge$ relation between the states of a parent and its child.

Any two caches $c_1, c_2$ in which neither is an ancestor of the other will
have a lowest common ancestor $x$ such that $x$ is an ancestor of both the
caches and $x$ has two distinct children $x_1$ and $x_2$ such that they are the
ancestors of $c_1$ and $c_2$ respectively. For any address $a$,
$x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are compatible by the second part
of Invariant \ref{localCompatibility}. From this, it straightaway follows that
$x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from Invariant
\ref{conservative}). From the previous paragraph, it follows that $c_1$ and
$c_2$ are also compatible. This proves Invariant \textit{nonAncestorCompatibility}.

We can prove that Invariant \textit{latestValue} is obeyed by strong induction
on time. Consider two cases: cache $c$ is already \clean{} for an address $a$
at time $t$ or the contrary.  If \clean{}, then it has the latest value at time
$t$. Since none of its children's directory entries can be in \Mo{}, all its
descendents should be no greater than \Sh{} for that address (as discussed
above). Any non-ancestor of $c$, and non-descendent of $c$ must be compatible
(as discussed above). Since $c$ is \clean, it is at least in the \Sh{} state,
so compatibility means that non-descendents can not be in \Mo{} state. All of
$c$'s ancestors are non leaves, so they can not process any store request. So,
no cache could have processed a store request other than $c$ itself, which
means that $c$ contains the latest value for address $a$ at time $t+1$ if it
was \clean{} at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child.  As can be seen in Figure \ref{trans}, whenever data is
sent either from a parent to a child, or from a child to the parent, the sender
is \clean. By the induction hypothesis, it means that the sender has the latest
value for the address. Because of Invariant \ref{noTransitWrite}, no store
request can be processed when the data is in transit, which means $c$ will have
the latest value for address $a$ at time $t+1$.

\subsection{Main intuition to prove the pair-wise properties}
We present two basic invariants that the MSI protocol of Figure \ref{trans}
obeys. In fact the protocol (\ie the request/response message semantics) was
designed to ensure the following two invariants. We believe that most
cache-coherence protocols would end up obeying these two invariants; we show
how MOSI, for instance, obeys these invariants in Section
\ref{sec:Refinements}. These two invariants together can prove all the pair-wise
properties of Section \ref{pairwise}.

@@

The MSI-based protocol in Figure \ref{trans} is such that any change in either
the state of a cache or its directory entry takes place only on sending or
receiving a response. The parent upgrades the corresponding directory entry on
sending a response to a child and the child downgrades its directory entry on
sending a response to the parent.

@@

>>>>

In this section, we present the proof structure for the soundness
property of our MSI-based cache coherence protocol of
Figure~\ref{trans}. Specifically, we formally reduce the proof of the
store atomicity theorem for the whole system to the proof of a set of
simple invariants by leveraging the ``intuitive'' justification of
correctness of our protocol.

%% This intuition
%% is not restricted to the MSI-based protocol;
%% Section~\ref{sec:Refinements} shows that it applies can apply to other
%% protocols.

@@

The store atomicity is a whole system property. Intuitively our
protocol is correct because it guarantees that at most one node can
write at a time, and that any local cache that attempts to read has
the ``latest'' value. We can state this formally with the following 4
invariants about nodes:

\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the
state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request has been
  processed at any time $t_i$ such that $t_i < t$, or
\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

We use the definition of compatibility between two caches as follows:
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible if and only if
either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
\label{compatible}
\end{defn}

@@

Given these invariants, we codify our intuition by showing that
Theorem~\textit{storeAtomicity} follows naturally from these
invariants.  A store may happen only when leaf cache is in \Mo{} state
for that address (Invariant~\textit{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\textit{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\textit{processReq}), showing Definition~\textit{uniqStTimes} holds.

Given Invariant~\textit{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\textit{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\textit{processReq}). As leaf caches have no children, a leaf cache in
at least in \Sh{} state for address $a$ is \clean for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\textit{lastestValue}, contains the ``latest
value''. Thus, the final condition for Theorem~\textit{storeAtomicity}
is also satisfied. All that remains is to show that these properties hold.

@@

We can directly show that Invariants~\textit{processReq} and
\textit{noReqAgain} hold at the initial configuration and are
preserved through each atomic transition of our protocol
(Figure~\ref{trans}). Invariants~\textit{uniqM} and
\textit{latestValue} correctness rely on properties of the node
hierarchy. We can prove both using the following invariants on
parent-child relationships.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \textit{conservative} and
\textit{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \textit{conservative} and the first part
\textit{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

Any two caches $c_1, c_2$ neither of which are an ancestor of the
other will have a lowest common ancestor $x$ which is an ancestor of
both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
such that they are the ancestors of $c_1$ and $c_2$ respectively. For
any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
compatible by the second part of
Invariant~\ref{localCompatibility}. From this, it straightaway follows
that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
Invariant~\ref{conservative}). From the previous paragraph, it follows
that $c_1$ and $c_2$ are also compatible. This proves
Invariant \textit{nonAncestorCompatibility}.

We prove Invariant \textit{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants can not be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they can not process any store
request. Therefore, no cache save $c$ may have process a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

If $c$ was not clean at time $t$, it must have received data from
either its parent or its child.  As can be seen in Figure~\ref{trans},
whenever data is sent either from a parent to a child, or from a child
to the parent, the sender is \clean. By the induction hypothesis, it
means that the sender has the latest value for the address. Because of
Invariant~\ref{noTransitWrite}, no store request can be processed when
the data is in transit, which means $c$ will have the latest value for
address $a$ at time $t+1$.

\subsection{Proving parent-child hierarchy properties}

%ndave: This needs some repointing

We present two basic invariants that the MSI protocol of
Figure~\ref{trans} obeys. In fact the protocol (\ie{} the
request/response message semantics) was designed to ensure the
following two invariants. We believe that most cache-coherence
protocols would end up obeying these two invariants; we show how MOSI,
for instance, obeys these invariants in
Section~\ref{sec:Refinements}. These two invariants together can prove
all the pair-wise properties of Section~\ref{pairwise}.

@@

The MSI-based protocol in Figure~\ref{trans} is such that any change in either
the state of a cache or its directory entry takes place only on sending or
receiving a response. The parent upgrades the corresponding directory entry on
sending a response to a child and the child downgrades its directory entry on
sending a response to the parent.

@@

%%  LocalWords:  Invariants invariants
